## Severe weather events and related economic consequences across USA



## Synopsis
Several weather events can case loss of life, injuries, significant property damage, and/or disruption to commerce. This report using NOAA Storm Data to summarize the top 10 harmful events related to population health and economic consequences due to the events.


## Data Processing
### Loading the data
The data was loaded from the [file repository](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2) in the Reproducibal Research Coursera Course. The data was in the form of a comma-separated-value file compressed via the bzip2 algorithm.

```{r loading data, echo=TRUE, cache = TRUE}
setInternet2(TRUE)
temp <- tempfile()
download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2", temp) 
StormData <- read.csv(temp, stringsAsFactor = FALSE)
unlink(temp)
```

The file was downloaded, extracted and then read into a data frame named as StormData uisng the above codes.

### Clean the data 
Event Type: All values are transformed into captial value and trimmed the leading and trailing whitespace.
Begin Date: remove the time zero and format to date type
Year of event: generate the year of event from begin date
Health Issues: generate the health problems due to the events, including injuries and fatalities. 
Cost: sum the property damage and crop damage DMG with EXP.


```{r, echo = TRUE}
library(stirngr)
StormData$EVTYPE <- toupper(str_trim(StormData$EVTYPE))

StormData$BGN_DATE <- gsub(" 0:00:00", "", StormData$BGN_DATE)
StormData$BGN_DATE <- as.Date(StormData$BGN_DATE, "%m/%d/%Y")

StormData$YearEvent <- as.numeric(format(StormData$BGN_DATE, "%Y"))

StormData$HealthEvent <- with(StormData, INJURIES+FATALITIES)

exp_transform <- function(e) {
    if (e %in% c("b", "B"))
        return(9)
    else if (e %in% c("m", "M"))
        return(6)
    else if (e %in% c("k", "K"))
        return(3)
    else if (e %in% c("h", "H"))
        return(2)
    else {
        return(0)
    }
} 
StormData$ProductCost <- StormData$PROPDMG * 10 ** sapply(StormData$PROPDMGEXP, FUN = exp_transform)
StormData$CropCost <- StormData$CROPDMG * 10 ** sapply(StormData$CROPDMGEXP, FUN = exp_transform)
StormData$ProductCost <- format(StormData$ProductCost, scientific = FALSE)
StormData$CropCost <- format(StormData$CropCost, scientific = FALSE)
StormData$Cost <- with(StormData, ProductCost + CropCost)
```

### Limit the year of data analysis
Due to a lack of good records in the earlier years, some records were removed to keep the record consistency. A bar chart was created to find the start year for the data analysis.
From the chart, it looks like an steady increase in events since 1995. We limited data for analysis since 1995.

```{r, echo = TRUE}
RecordsByYear <- data.frame(table(StormData$YearEvent))
plot(RecordsByYear$Var1, RecordsByYear$Freq, xlab = "Year", ylab = "Total Events", 
     main = "Number of storm events by year", type = "l")
SubsetData <- subset(StormData, StormData$YearEvent >= 1995)
```


